HO unification from object language to meta language

### 2024-03-27 paper/main.tex
update tex

\noindent
We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
the one of \Ho.
Both have the invariant that they are not cyclic, TODO explain.
Other invariant: the terms in ho\_subst never contains eta and beta expansion

@@

Decompiling the substitution requires to first force the
progress of links and then allocating new unassigned variables
in the substitution for \Fo{} and finally decompiling all
assignments. Note that \ref{inv:linklhs} and the
occur check allows us to update the subst.

@@

\noindent
Decompiling an assignment requires to turn abstractions into
lambdas. For aestetic purposes we also eta-contract the result
(not needed since Fo equality can do that)

@@

\begin{gather}
\lambda x y.F~y~x = \lambda x y.x\\
\lambda x. f~(F~x)~x = f~(\lambda y.y)
\end{gather}
Note that here $F$ is used with different arities, moreover
in the second problem the left hand side happens to be an
eta expansion (of $f (\lambda y.y)$) only after we discover that
$F = \lambda x\lambda y.y$ (i.e. that $F$ discards the $x$ argument).
Both problems are addressed in the next section.

@@

\linketa are meant to suspend the unification of two terms, that is, if we have
a \maybeeta term $t$ which should be unified with a term $t'$ we don't want to
unify them with \Ue. As said before this would break unification but it can also
introduce \maybeeta in the substitution of \Ho.

In order to activate a \linketa, we need to implement new rules for the
\elpiIn{progress1} predicate. There are two cases making a \linketa to progress,
1) \lhs is instantiated to a rigid term, in this case \lhs is unified with \rhs.
TODO: the right eta hand side is eta-expanbded if it is an app/con; 2) \rhs can
be $\eta$-reduced to a term with rigid head, in this case \lhs and \rhs are
again unified. If one of these two condition is satisfied, the link has
fulfilled is task and can be removed from the list of suspended links; if none
the condition succeeds, the link is kept for a further iteration of
\elpiIn{progress}.

TODO: dire che scendiamo sotto i vari abs che formano il contesto? 
TODO: example for case 1: $\lambda x. \lambda y. F \appsep y \appsep x = f$
TODO: example for case 2: $\lambda x. \lambda y. F \appsep y \appsep x = G, F = \lambda x. \lambda y. a$

A second way to progress \linketa, that we call \linketa deduplication, is to
check if \linkStore contains two \linketa $l_1$ and $l_2$ with same \lhs. This
situation occurs if two \maybeeta terms are unified with a same unification
variable. In this case, we can unify the $l_1$ and $l_2$ \rhs (that, by
construction are both on the form $\lambda x. T_x$) and remove one of the two
links.

@@

We can note that the 1) insertion of these rules for \elpiIn{progress1} do not
prevent the termination of \elpiIn{progress}, since, a link activation runs
terminating operations (such as term unification and link-removal); 2) \linketa
deduplication runs again terminating operations and 3) if none of these two 
situations is performed, then the substitution and \linkStore
Note that if the link remain suspended \elpiIn{progress} continues to terminate,
since, it would mean that the condition of the \elpiIn{if}-branchement succeds
and the \elpiIn{progress} terminates its execution.

\begin{lemma}
  We never add eta-expansions in the substitution
\end{lemma}

\begin{proof}[Proof sketch]
\end{proof}

TODO: we can however have $\lambda x. F_x$ if we know that $F$ does not reduce
to $T x$ where x is not free in $T$.

>>>>

\noindent
We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
the one of \Ho.
Both have the invariant that they are not cyclic, TODO: explain.

\begin{invariant}
  Any term in the substitution of \Ho is in normal form.
  TODO: move this to the right place
  TODO: definire normal form, maybe in definition 2.7
  \label{inv:subst-nf}
\end{invariant}
\begin{invariant}

  Any variable in the substitution with arity \elpiIn{N} is always fully applied
  TODO: move this to the right place
  TODO: definire normal form, maybe in definition 2.7
  \label{inv:fully-applied-vars}
\end{invariant}

@@

Decompiling the substitution requires to first force the
progress of links and then allocating new unassigned variables
in the substitution for \Fo{} and finally decompiling all
assignments. Note that \cref{inv:linklhs} and the
occur check allows us to update the subst.

@@

TODO: What is commit-links and complete-mapping?, maybe complete-mapping can be
hidden in the code rendering?

\noindent
Decompiling an assignment requires to turn abstractions into
lambdas. For aestetic purposes we also eta-contract the result
(not needed since \Fo{} equality can do that)

@@

\begin{gather}
\lambda x y.F \appsep y \appsep x = \lambda x y.x \label{eq:unif-eta1}\\
\lambda x. f \appsep (F \appsep x) \appsep x = f \appsep (\lambda y.y) \label{eq:unif-eta2}
\end{gather}
Note that here $F$ is used with different arities, moreover
in the second problem the left hand side happens to be an
eta expansion (of $f (\lambda y.y)$) only after we discover that
$F = \lambda x\lambda y.y$ (i.e. that $F$ discards the $x$ argument).
Both problems are addressed in the next section.

@@

\linketa are meant to delay the unification of ``problematic'' terms. In the
following, we call \linkStore the list of suspended links.

In order to activate a \linketa $l$, we need to implement new rules for the
\elpiIn{progress1} predicate. After passing under all the \elpiIn{abs}
constructors of $l$, there are two cases making a \linketa to progress, 1) \lhs
is instantiated to a rigid term 2) \rhs can be $\eta$-reduced to a term with
rigid head. If \lhs is instantiated to a rigid term $t$, by \cref{inv:subst-nf},
we know that $t$ does not contain any \maybeeta. Let $t'$ the right hand side,
if $t$ is a constant or a function application, then, $t'$, which by
construction has \elpiIn{lam} as head, should be an $\eta$-expansion. We are
therefore allowed to unify $\lambda x. t \appsep x$ (the $\eta$-expanded version
of $t$) with $t'$. Finally, if \lhs starts has \elpiIn{lam} as constructor, then
$t$ is not an $\eta$-expansion and therefore, $t$ can be unified with $t'$.

The second way to activate a \linketa is when the \rhs can $\eta$-reduced to a
term $t$ with rigid head, i.e. $t'$ is not a \maybeeta. This means that we can
unify \lhs with $t'$.

Once a \linketa is activated, it can be removed from \linkStore,
otherwise, the link is kept for a further iteration of \elpiIn{progress}. Note
that this link progression enforce
\cref{inv:subst-nf,inv:linklhs,inv:link-eta-right}: we never commit a term in
the \Ho substitution, since we make unification only when we know that the 
terms are no more \maybeeta, and when \lhs is no more a variable or \rhs is 
no more a \maybeeta, the link is removed from \linkStore.

TODO: example for case 1: $\lambda x. \lambda y. F \appsep y \appsep x = f$
TODO: example for case 2: $\lambda x. \lambda y. F \appsep y \appsep x = G, F = \lambda x. \lambda y. a$

A second way to progress \linketa, that we call \linketa deduplication, is when
\linkStore contains two \linketa $l_1$ and $l_2$ with a \lhs having the same
variable address. Let the \lhs of $l_1$ be \elpiIn{uva U R} and the \lhs of
$l_2$ be \elpiIn{uva V S}, then, by \cref{inv:fully-applied-vars}, \elpiIn{R}
and \elpiIn{S} have same scope. Let $t$ be the term obtained by replacing all
each name \elpiIn{S}$_i$ in the \rhs of $l_1$ with \elpiIn{R}$_i$, $t$ is
unified with the \rhs of $l_2$ and one of the two links between $l_1$ and $l_2$
is removed from \linkStore.

@@

We can note that the insertion of these rules for \elpiIn{progress1} do not
prevent the termination of \elpiIn{progress}, since: if a link is activated it
is removed from \linkStore and the recursive call to \elpiIn{progress} will have
a smaller list of links to recurse on. Moreover, link activation only runs
unification which are terminating instructions. If a link is deduplicated, the
termination of \elpiIn{progress} is still guaranteed since again we reduce
\linkStore and the instructions run by link deduplications are all terminating.
Finally, if a link is neither activated nor deduplicated, i.e. it remains
suspended, then the condition of the \elpiIn{if}-statement of \elpiIn{progress}
succeeds making it to terminate.

TODO: we can have $\lambda x. F_x$ in the substitution if we know that $F$ does
not reduce to $T x$ where x is not free in $T$.

\subsection{Example}

In this subsection we show two complete runs of the extended version of \Ue.

\paragraph{Example 1}
\begin{center}
  \elpiIn{flam x\ flam y\ fapp[fuva A, y, x] = flam x\ flam y\ x}
\end{center}
% $$\flambda xy.F \appsep y \appsep x = \lambda xy.x$$
\begin{itemize}
  \item After compilation: terms are $X0 = \lambda c0.\lambda c1.c0$ with links
        $\linketaM{c0}{X1_{c0}}{\lambda c1.X2_{c1~c0}}, \linketaM{}{X0}{\lambda
        c0.X1_{c0}}$ and mappings: $\mapping{C0}{X2}{2}$
  \item After unification $X0$ is set to $\lambda c0.\lambda c1.c0$
  \item This unification wakes up the second link, since its \lhs is now
        instantiated. $X0$ is unified with its \rhs, which assigns $X1_{c0} :=
        \lambda c1.c0$
  \item The instantiation of $X1$ wakes the first link, and $\lambda c1.c0$ is
        unified with $\lambda c1.X2_{c1~c0}$, setting $X2 _{c1~c2} := c2$ by
        pruning.
  \item All the links have been solved and the decompilation, assigns the \Fo{}
        variable \elpiIn{A} to $\lambda xy.y$ 
\end{itemize}

\paragraph{Example 2}
\begin{center}
  \elpiIn{flam x\ fapp[f, fapp[fuva F, x], x] = fapp[f, flam x\ x]}
\end{center}
% $$\flambda xy.F \appsep y \appsep x = \lambda xy.x$$
\begin{itemize}
  \item After compilation: terms are $X0 = f~\lambda c0.c0$ with links
        $\linketaM{}{X0}{\lambda c0.(f~X1_{c0}~c0)}$ and mappings:
        $\mapping{C0}{X1}{1}$
  \item After unification $X0$ is set to $f~\lambda c0.c0$
  \item This unification wakes up the link, since its \lhs is now
        instantiated. $X0$ is unified with its \rhs, which assigns $X1_{c0} :=
        \lambda c1.c0$
  \item The instantiation of $X1$ wakes the first link, and $\lambda c1.c0$ is
        unified with $\lambda c1.X2_{c1~c0}$, setting $X2 _{c1~c2} := c2$ by
        pruning.
  \item All the links have been solved and the decompilation, assigns the \Fo{}
        variable \elpiIn{A} to $\lambda xy.y$ 
\end{itemize}
